---
title: "WINGMEN - TIMER Plots"
author: "Fiona Hartley"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: 
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)

```

# Set directories
Set the data directory as the folder where you have stored the data I've sent to you. It's probably easier to keep all of these in one place then you don't have to keep changing the code. I use the file names I gave to each of the objects to load them in, so if you changed the file names of anything I sent to you, you'll have to go through and change those within the code too.

You should set the results directory to a folder where you want to store any plots produced.

```{r set_directories}

data_directory <- "C:/Users/fhartley/Desktop/data"
res_directory <- "C:/Users/fhartley/Desktop/plots"

```

# Load packages

```{r load_packages, echo=FALSE, message=FALSE, warning=FALSE}

# Annotation
library(org.Hs.eg.db)

# Graphics
library(ggplot2)
library(plotly)

# Other
library(dplyr)
library(tidyverse)

```

# Load the TIMER results

```{r load_files}

# Read in the TIMER results
timer <- readRDS(file = paste(data_directory, "TIMER.rds", sep = "/"))

# Read in the metadata
metadata <- readRDS(file = paste(data_directory, "metadata.rds", sep = "/"))
metadata <- metadata[metadata$WINGMEN.sampleID %in% timer$ID,]

```

# Prepare the data
You don't need to change anything here

```{r prepare_input}

if(all(timer$ID == metadata$WINGMEN.sampleID)){
  timer$WINGMEN.patientID <- metadata$WINGMEN.patientID
  timer$WINGMEN.sampleID <- metadata$WINGMEN.sampleID
}

# Tidy up the column names
colnames(timer)[2] <- "B Cells"
colnames(timer)[3] <- "CD4+ T Cells"
colnames(timer)[4] <- "CD8+ T Cells"
colnames(timer)[5] <- "Neutrophils"
colnames(timer)[6] <- "Macrophages"
colnames(timer)[7] <- "Dendritic Cells"

```

# Cell type results
Create a vector of cell types you'd like to view the results for 
This can be any length from one cell type or the complete set.

```{r chosen_cell_types}

# Possible cell types
colnames(timer)[2:7]

# You can type your cell types of interest using the format below 
my_cells <- colnames(timer)[2:7]

```

## Create plots of the TIMER scores for each cell type with individual patients' PRE and POST samples connected. 
Lots of this code just changes the way the plot looks. You can play around with colours and sizes to make it look how you want it to.

```{r cell_type_plots}

#c("#F8766D", "#00BFC4")
p1 <- list()

for(i in my_cells){
  # Title
  my_title <- i
  
  # Generate the initial ggplots
  p1[[i]] <- ggplot(timer, 
                    aes(Treatment, get(i), fill = Treatment, text = WINGMEN.sampleID)) + 
             geom_boxplot()+ 
             geom_point()+
             ylab(my_title)+
             scale_fill_manual(values = c("black", "#f89b7c"))+ # You can change the colours of the PRE and POST boxplots
              theme(
                legend.position = "none",
                panel.border = element_blank(),
                panel.background = element_rect(fill = "transparent"),
                plot.background = element_rect(fill ="transparent", color = NA),
                axis.line.x = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
                axis.line.y = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
                axis.title.x = element_text(colour = "black", size = 12),
                axis.title.y = element_text(colour = "black", size = 12),
                axis.text.y = element_text(color = "black", size = 10),
                axis.text.x = element_text(color = "black", size = 10),
                plot.title = element_text(hjust = 0.5))+
             ggtitle(i)
  
  # Add a geom_segment line for each patient pair to connect the dots
  for(patient in metadata$WINGMEN.patientID[duplicated(metadata$WINGMEN.patientID)]){
      p1[[i]] <- p1[[i]]+geom_segment(data = timer[timer$WINGMEN.patientID == patient,], 
                                      aes(x = as.numeric(Treatment[Treatment == "PRE"]),
                                          y = get(i)[2],
                                          xend = as.numeric(Treatment[Treatment == "POST"]),
                                          yend = get(i)[1]))
  }

  
  # Make the plots interactive
  p1[[i]] <- ggplotly(p1[[i]], tooltip = "text")
  
  # Calculate the significance
  temp.df <- timer[, c("ID", i)]
  
  temp.df1 <- temp.df[grep("PRE", temp.df$ID),]
  temp.df1$ID <- as.numeric(gsub("PRE", "", temp.df1$ID))
  colnames(temp.df1)[2] <- "PRE"
  
  temp.df2 <- temp.df[grep("POST", temp.df$ID),]
  temp.df2$ID <- as.numeric(gsub("POST", "", temp.df2$ID))
  colnames(temp.df2)[2] <- "POST"
  
  temp.df <- merge(temp.df1, temp.df2, by = "ID", all = T)
  
  # Add this value to the interactive plot
  p1[[i]] <- p1[[i]] %>% 
                add_annotations(
                  text = paste("p =", round(wilcox.test(as.numeric(temp.df$PRE), temp.df$POST, paired = T)$p.value, 3)), 
                  x = 1.5,  
                  y = 1,
                  xref = "x",
                  yref = "paper",
                  showarrow = FALSE,  
                  font = list(size = 14, color = "black")
                )
}

# View the plot(s)
p1


# Save a plot. Here you will need to change the cell in the p1$`Resident CD8 memory cell` to your cell type of interest, and the filename. Once you have run the code, it will appear in the viewer. You can then download the plot from the options menu that appears above the plot. Unfortunately I haven't found a way to automate this!
plotly::config(p1$`CD8+ T Cells`,
    toImageButtonOptions = list(
      format = "png", # one of png, svg, jpeg, webp
      filename = "CD8 T Cells",
      width = 300,
      height = 350,
      scale = 5
    ))


```

## Alternatively, if making plots in GraphPad is easier, this code will extract the scores counts for your chosen cell types and export it in a format that is easy to copy and paste over. 
A separate csv will be created for each cell type.

```{r cell_type_tables}

t1 <- list()

for(i in my_cells){
  data <- timer
  data <- data[,c("ID", i)]
  rownames(data) <- data$ID

  pre <- data[grep("PRE", rownames(data)),]
  pre$WINGMEN.patientID <- gsub("PRE", "", pre$ID)
  colnames(pre)[2] <- "PRE"
  
  post <- data[grep("POST", rownames(data)),]
  post$WINGMEN.patientID <- gsub("POST", "", post$ID)
  colnames(post)[2] <- "POST"

  data <- merge(pre[,c(2,3)], 
                post[,c(2,3)], 
                by = "WINGMEN.patientID", 
                all = T)
  
  data$WINGMEN.patientID <- as.numeric(data$WINGMEN.patientID)
  
  t1[[i]] <- arrange(data, WINGMEN.patientID)

  }

# Save the tables. A different csv file will be created for cell type gene
for(i in names(t1)){
  write.csv(t1[[i]], paste0(res_directory, "/", i, ".csv"))
}

```

# Immune signature results
Create a vector of immune signatures you'd like to view the results for 
This can be any length from one to the complete set.

```{r chosen_immune_signatures}

# Possible cell types
colnames(immune_sigs)

# You can type your cell types of interest using the format below 
my_immune_sigs <- c("CD8_Palmer",
                    "IFNG.GS",
                    "ISG.RS")

```

## Create plots of the ssGSEA enricment scores for each cell type with individual patients' PRE and POST samples connected. 
Lots of this code just changes the way the plot looks. You can play around with colours and sizes to make it look how you want it to.

```{r immune_sig_plots}

p2 <- list()
for(i in my_immune_sigs){
  # Title
  my_title <- gsub("_", " ", i)
  my_title <- paste0(my_title, " Score")
  
  # Generate the initial ggplots
  p2[[i]] <- ggplot(immune_sigs, 
                    aes(Treatment, get(i), fill = Treatment, text = WINGMEN.sampleID)) + 
             geom_boxplot()+ 
             geom_point()+
             ylab(my_title)+
             #ggtitle("Your title")+
             theme(
                legend.position = "none",
                panel.border = element_blank(),
                panel.background = element_rect(fill = "transparent"),
                plot.background = element_rect(fill ="transparent", color = NA),
                axis.line.x = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
                axis.line.y = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
                axis.title.x = element_text(colour = "black", size = 12),
                axis.title.y = element_text(colour = "black", size = 12),
                axis.text.y = element_text(color = "black", size = 10),
                axis.text.x = element_text(color = "black", size = 10),
                plot.title = element_text(hjust = 0.5))
  
  # Add a geom_segment line for each patient pair to connect the dots
  for(patient in metadata$WINGMEN.patientID[duplicated(metadata$WINGMEN.patientID)]){
      p2[[i]] <- p2[[i]]+geom_segment(data = immune_sigs[immune_sigs$WINGMEN.patientID == patient,], 
                                      aes(x = as.numeric(Treatment[Treatment == "PRE"]),
                                          y = get(i)[2],
                                          xend = as.numeric(Treatment[Treatment == "POST"]),
                                          yend = get(i)[1]))
  }
  
  # Make the plots interactive
  p2[[i]] <- ggplotly(p2[[i]], tooltip = "text")
  
  # Calculate the significance
  temp.df <- immune_sigs[, c("WINGMEN.sampleID", i)]
  
  temp.df1 <- temp.df[grep("PRE", temp.df$WINGMEN.sampleID),]
  temp.df1$WINGMEN.sampleID <- as.numeric(gsub("PRE", "", temp.df1$WINGMEN.sampleID))
  colnames(temp.df1)[2] <- "PRE"
  
  temp.df2 <- temp.df[grep("POST", temp.df$WINGMEN.sampleID),]
  temp.df2$WINGMEN.sampleID <- as.numeric(gsub("POST", "", temp.df2$WINGMEN.sampleID))
  colnames(temp.df2)[2] <- "POST"
  
  temp.df <- merge(temp.df1, temp.df2, by = "WINGMEN.sampleID", all = T)
  
  significant <- wilcox.test(as.numeric(temp.df$PRE), temp.df$POST, paired = T)$p.value
  
  # Add this value to the interactive plot
  p2[[i]] <- p2[[i]] %>% 
                add_annotations(
                  text = paste("p =", round(wilcox.test(as.numeric(temp.df$PRE), temp.df$POST, paired = T)$p.value, 3)), 
                  x = 1.5,  
                  y = 1.05,
                  xref = "x",
                  yref = "paper",
                  showarrow = FALSE,  
                  font = list(size = 14, color = "black")
                )
}

# View the plot(s)
p2


# Save a plot
plotly::config(p2$ISG.RS,
    toImageButtonOptions = list(
      format = "png", # one of png, svg, jpeg, webp
      filename = "ISG.RS",
      width = 300,
      height = 350
    ))


```

Alternatively, if making plots in GraphPad is easier, this code will extract the scores counts for your chosen cell types and export it in a format that is easy to copy and paste over. A separate csv will be created for each cell type.

```{r immune_sig_tables}

t2 <- list()
for(i in my_immune_sigs){
  data <- immune_sigs
  data$ID <- rownames(data)
  data <- data[,c("ID", i)]

  pre <- data[grep("PRE", rownames(data)),]
  pre$WINGMEN.patientID <- gsub("PRE", "", pre$ID)
  colnames(pre)[2] <- "PRE"
  
  post <- data[grep("POST", rownames(data)),]
  post$WINGMEN.patientID <- gsub("POST", "", post$ID)
  colnames(post)[2] <- "POST"

  data <- merge(pre[,c(2,3)], 
                post[,c(2,3)], 
                by = "WINGMEN.patientID", 
                all = T)
  
  data$WINGMEN.patientID <- as.numeric(data$WINGMEN.patientID)
  
  t2[[i]] <- arrange(data, WINGMEN.patientID)

  }

# Save the tables. A different csv file will be created for each gene
for(i in names(t2)){
  write.csv(t2[[i]], paste0(res_directory, "/", i, ".csv"))
}

```
